Examining background gene lists in clusterProfiler - RNA-seq example

### 2024-06-05 analysis/analyse_rnaseq1.Rmd
update figs

compare_ora <- function(bad,good){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  mup <- merge(badup,goodup,by=0)
  mdn <- merge(baddn,gooddn,by=0)
  bup <- subset(mup,p.adjust.x<0.05)$Row.names
  gup <- subset(mup,p.adjust.y<0.05)$Row.names
  bdn <- subset(mdn,p.adjust.x<0.05)$Row.names
  gdn <- subset(mdn,p.adjust.y<0.05)$Row.names
  sig_bad <- unique(c(bup,bdn))
  sig_good <- unique(c(gup,gdn))
  nsig_bad <- length(sig_bad)
  nsig_good <- length(sig_good)
  jac <- length(intersect(sig_bad,sig_good)) / length(union(sig_bad,sig_good))
  result=c("nsig_bad"=nsig_bad,"nsig_good"=nsig_good,"jaccard"=jac)
  return(result)
}

>>>>

compare_ora <- function(bad,good){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  mup <- merge(badup,goodup,by=0)
  mdn <- merge(baddn,gooddn,by=0)
  bup <- subset(mup,p.adjust.x<0.05)$Row.names
  gup <- subset(mup,p.adjust.y<0.05)$Row.names
  bdn <- subset(mdn,p.adjust.x<0.05)$Row.names
  gdn <- subset(mdn,p.adjust.y<0.05)$Row.names
  sig_bad <- unique(c(bup,bdn))
  sig_good <- unique(c(gup,gdn))
  nsig_bad <- length(sig_bad)
  nsig_good <- length(sig_good)
  jac <- length(intersect(sig_bad,sig_good)) / length(union(sig_bad,sig_good))
  result=c("nsig_bad"=nsig_bad,"nsig_good"=nsig_good,"Jaccard"=jac)
  return(result)
}

### 2024-06-05 analysis/analyse_rnaseq2.Rmd
update figs

The goal of this work is to understand how ClusterProfiler manages the background list,
as we have observed some weird behaviour.
In particular we see that when we provide a custom background, those genes do not appear in the
final analysis unless they are also included in the gene list.

In the code chunk below called `libs`, you can add and remove required R library dependancies.
Check that the libraries listed here match the Dockerfile, otherwise you might get errors.

@@

# bad=ORA analysis, a list with results of up and down regulated pathways
# good=same as bad, except fixed the background problem
compare_ora_plots <- function(bad,good){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  mup <- merge(badup,goodup,by=0)

  MAX=max(-log10(c(mup$p.adjust.x,mup$p.adjust.y)))

  plot(-log10(mup$p.adjust.x), -log10(mup$p.adjust.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="-log10 FDR")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

@@

  plot(-log10(mdn$p.adjust.x), -log10(mdn$p.adjust.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="-log10 FDR")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)
}

compare_ora <- function(bad,good){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  bup <- rownames(subset(badup,p.adjust<0.05))
  gup <- rownames(subset(goodup,p.adjust<0.05))
  bdn <- rownames(subset(baddn,p.adjust<0.05))
  gdn <- rownames(subset(gooddn,p.adjust<0.05))
  sig_bad <- unique(c(bup,bdn))
  sig_good <- unique(c(gup,gdn))
  nsig_bad <- length(sig_bad)
  nsig_good <- length(sig_good)
  jac <- length(intersect(sig_bad,sig_good)) / length(union(sig_bad,sig_good))
  result=c("nsig_bad"=nsig_bad,"nsig_good"=nsig_good,"jaccard"=jac)
  return(result)
}

@@

summary(res1$nsig_bad)
summary(res1$nsig_good)
summary(res1$jaccard)

@@

par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
barplot(t(df),beside=TRUE,horiz=TRUE,las=1,
  xlab="no. significant sets", main="Comparison of original and corrected data",
  col=c("gray30","gray80"))

legend("bottomright", inset=.02, legend=c("original","corrected"),
  fill=c("gray30","gray80"), horiz=TRUE, cex=1)

@@

barplot(sort(rat2),horiz=TRUE,las=1,main="increase in significant sets after correction",xlab="fold increase")

# JACCARD
jac <- matrix( res1$jaccard , ncol=9 )
rownames(jac) <- names(d)
colnames(jac) <- names(gs)

@@

lapply(ngenes, function(n) {
  res2 <- subset(res1,Var3==n) ; summary(res2$jaccard)
} )

@@

# JACCARD
jac <- lapply(ngenes ,function(n) {
  res2 <- subset(res1,Var3==n)
  ns <- matrix( res2$jaccard , ncol=9 )
  rownames(ns) <- names(d)
  colnames(ns) <- names(gs)
  return(ns)
} )
names(jac) <- ngenes

@@

plot(jac[,"c5.go"],type="b", pch=19,cex=1.5,lwd=2,
  xaxt = "n",ylim=c(0,1),ylab="jaccard",xlab="no. genes selected")
axis(1, at = seq(1, 5, by = 1),
     labels = ngenes)
points(jac[,"c2.cp.reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
points(jac[,"c8.all"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
points(jac[,"c3.tft.gtrd"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
mtext("impact of gene list length on severity of FDR problem")

legend("bottomright", legend=c("GO","Reactome", "Markers", "TFT"),
       col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)

@@

```{r,charts1}

@@

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=125)
  good <- oragood(d[[1]],gs[[i]],ngenes=125)
  compare_ora_plots(bad,good)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=250)
  good <- oragood(d[[1]],gs[[i]],ngenes=250)
  compare_ora_plots(bad,good)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=500)
  good <- oragood(d[[1]],gs[[i]],ngenes=500)
  compare_ora_plots(bad,good)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=1000)
  good <- oragood(d[[1]],gs[[i]],ngenes=1000)
  compare_ora_plots(bad,good)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=2000)
  good <- oragood(d[[1]],gs[[i]],ngenes=2000)
  compare_ora_plots(bad,good)
} )

>>>>

The goal of this work is to understand the behaviour of ClusterProfiler's p-value adjustment method
as we have observed some weird behaviour.
In particular we see that clusterprofiler's criteria for inclusion/exclusion of gene sets is irregular.
For example clusterprofiler seems to include gene sets as detected if there is just 1 member detected
in the foreground list.
If a gene set has 10 in the background and  0 in the foreground it looks like it gets excluded.
This is a bit strange and contrary to best practice, where the detection threshold should be set by
the overlap size in the background list, rather than the foreground list.

Theoretically, this behaviour could have a negative impact on performance, as FDR is applied to the
wrong pathways.

@@

# bad=ORA analysis, a list with results of up and down regulated pathways
# good=same as bad, except fixed the background problem
compare_ora_plots <- function(bad,good,libname=NULL,setsize=NULL){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  mup <- merge(badup,goodup,by=0)

  MAX=max(-log10(c(mup$p.adjust.x,mup$p.adjust.y)))
  UPHEADER=paste(libname,"up",setsize,"genes")
  DNHEADER=paste(libname,"dn",setsize,"genes")

  par(mfrow=c(1,2))

  plot(-log10(mup$p.adjust.x), -log10(mup$p.adjust.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original -log10 FDR", ylab="corrected -log10 FDR", main=UPHEADER)
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

@@

  plot(-log10(mdn$p.adjust.x), -log10(mdn$p.adjust.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original -log10 FDR", ylab="corrected -log10 FDR", main=DNHEADER)
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)
}

compare_ora <- function(bad,good){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  bup <- rownames(subset(badup,p.adjust<0.05))
  gup <- rownames(subset(goodup,p.adjust<0.05))
  bdn <- rownames(subset(baddn,p.adjust<0.05))
  gdn <- rownames(subset(gooddn,p.adjust<0.05))
  sig_bad <- unique(c(bup,bdn))
  sig_good <- unique(c(gup,gdn))
  nsig_bad <- length(sig_bad)
  nsig_good <- length(sig_good)
  jac <- length(intersect(sig_bad,sig_good)) / length(union(sig_bad,sig_good))
  result=c("nsig_bad"=nsig_bad,"nsig_good"=nsig_good,"Jaccard"=jac)
  return(result)
}

@@

summary(res1$nsig_bad)
summary(res1$nsig_good)
summary(res1$Jaccard)

@@

par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,330),
  xlab="no. significant sets", main="Effect of correcting the FDR bug",
  col=c("gray30","gray80"))
text(df$ns1+20,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
text(df$ns2+20,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
legend("bottomright", inset=.02, legend=c("corrected","original"),
  fill=c("gray80","gray30"), horiz=FALSE, cex=1)

@@

# JACCARD
jac <- matrix( res1$Jaccard , ncol=9 )
rownames(jac) <- names(d)
colnames(jac) <- names(gs)

@@

lapply(ngenes, function(n) {
  res2 <- subset(res1,Var3==n) ; summary(res2$Jaccard)
} )

@@

# JACCARD
jac <- lapply(ngenes ,function(n) {
  res2 <- subset(res1,Var3==n)
  ns <- matrix( res2$Jaccard , ncol=9 )
  rownames(ns) <- names(d)
  colnames(ns) <- names(gs)
  return(ns)
} )
names(jac) <- ngenes

@@

plot(jac[,"GO"],type="b", pch=19,cex=1.5,lwd=2,
  xaxt = "n",ylim=c(0,1),ylab="Jaccard index",xlab="no. genes selected")
axis(1, at = seq(1, 5, by = 1),
     labels = ngenes)
points(jac[,"Reactome"],type="b",col="red",pch=19,cex=1.5,lwd=2)
points(jac[,"Cellmarkers"],type="b",col="blue",pch=19,cex=1.5,lwd=2)
points(jac[,"TFT GTRD"],type="b",col="gray50",pch=19,cex=1.5,lwd=2)
mtext("impact of gene list length on FDR problem severity")

legend("bottomright", legend=c("GO","Reactome", "Cellmarkers", "TFT"),
       col=c("black", "red", "blue", "gray50"), lty=1, cex=1.2,lwd=3,pch=19)

@@

```{r,charts1,fig.height=4}

@@

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=125)
  good <- oragood(d[[1]],gs[[i]],ngenes=125)
  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=125)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=250)
  good <- oragood(d[[1]],gs[[i]],ngenes=250)
  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=250)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=500)
  good <- oragood(d[[1]],gs[[i]],ngenes=500)
  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=500)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=1000)
  good <- oragood(d[[1]],gs[[i]],ngenes=1000)
  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=1000)
} )

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]],ngenes=2000)
  good <- oragood(d[[1]],gs[[i]],ngenes=2000)
  compare_ora_plots(bad,good,libname=names(gs)[i],setsize=2000)
} )
